---
title: "SuMO-Fil: Supervised Multi-Omic Filtering"
author: "Lorin Towle-Miller"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(SuMOFil)
```

## Package Info

Provide functions for simulating gene pathway networks with a continuous outcome, and a filtering method to trim some of the high dimensional noise. Note that under this setting, it is assumed that data type G has pathway elements that relate to pathway elements in data type X, which then those pathway elements in X relate to the outcome Y. 

In addition to the pathway elements, there are elements within X that are related to Y but not G, elements within X that are related to G but not Y, and elements within X that are not related to G or Y Similarly, there are elements within G that are related to Y and not X, elements within G that are related to X and not Y, and elements within G that are no related to X or Y. 

The goal of Supervised Multi-Omic Filtering (SuMO-Fil) is to remove elements from X and G that are not pathway elements.

## Functions

### simPathwaySystem

simPathwaySystem(alpha, Rsq, nsample, rind, nx, nxPrime, ns, ntranscriptNoise, ng, ngPrime, nh, ngeneNoise, noiseSD)

**Purpose** 

Simulate a pathway network between two data types and a corresponding continuous outcome.

**Parameters**

- *alpha*: Factor that controls the differences in strength between the relationships the pathway elements of X and G as compared to the strength in the relationship between Y with X and G in the pathway.

- *Rsq*: Factor controlling the differences in variances across X as compared to X given G

- *nsample*: Number of observations to generate in the system

- *rind*: Base measure for the variance within an observation

- *nx*: Number of pathway elements in the data type X (related to G and Y)

- *nxPrime*: Number of elements within X that are related to G and not Y

- *ns*: Number of elements within X that are related to Y and not G

- *ntranscriptNoise*: Number of elements within X that are not related to G or Y

- *ng*: Number of pathway elements in the data type G (related to X and Y)

- *ngPrime*: Number of elements within G that are related to X and not Y

- *nh*: Number of elements within G that are related to Y and not X

- *ngeneNoise*: Number of elements within G that are not related to X or Y

- *noiseSD*: Standard deviation to use on the noise elements in both X and G

**Details**

All simulations are performed on a multivariate normal distribution and parameters are collected that control the covariance structure between all element types in X, all element types in G, and outcome Y.

**Value**

A list containing the following values:

- *x*: Dataframe of the simulated X values with dimensions nsample (rows) by (nx + nxPrime + ns + ntranscriptNoise) (columns)

- *g*: Dataframe of the simulated G values with dimensions nsample (rows) by (ng + ngPrime + nh + ngeneNoise) (columns)

- *y*: Dataframe of the simulated outcome Y with dimensions nsample (rows) by 1

- *indicatorX*: A vector of length (nx + nxPrime + ns + ntranscriptNoise) that labels the corresponding columns within x as either "pathway" or "not in pathway"

- *indicatorG*: A vector of length (ng + ngPrime + nh + ngeneNoise) that labels the corresponding columns within g as either "pathway" or "not in pathway"

**Examples**

```{r}
#Simulate a system with 200 observations
simSystem <- simPathwaySystem(
  alpha = 0.35
  ,Rsq = 0.35
  ,nsample = 200
  ,rind = 0.35
  ,nx = 15
  ,nxPrime = 100
  ,ns = 50
  ,ntranscriptNoise = 100
  ,ng = 10
  ,ngPrime = 100
  ,nh = 30
  ,ngeneNoise = 100
  ,noiseSD = 0.5
)
```

### SuMOFil

SuMOFil(x, g, y,numClusters_1, numClusters_2)

**Purpose**  

Remove elements from x and g that are not related to each other or y.

**Parameters**

- *x*: A matrix of the X values where the rows represent the samples and the columns represent the elements

- *g*: A matrix of the G values where the rows represent the samples and the columns represent the elements

- *y*: A vector (or dataframe with 1 column) of the continuous outcomes for each of the samples

- *numClusters_1*: Number of k-means clusters to use for filtering on the first round (between relationships of X and Y and relationships of G and Y)

- *numClusters_2*: Number of k-means clusters to use for filtering on the second round (between relationships of X and G)

**Details**

The goal of SuMO-Fil is to remove elements in X and G that are not related to eachother OR the outcome Y. By filtering out elements, it can speed up underlying pathway analyses.

The first round in the filtering process compares the relationships between X and Y and the relationships between G and Y. SuMO-Fil will find the absolute value of the correlations between each element of X with the outcome Y, and then perform k-means clustering the absolute correlations. All elements of X that correspond to the lowest cluster are noted (not yet filtered unless criteria from the second round is also met). A similar process is repeated for measuring the correlations between G and Y.

The second round in SuMO-Fil compares the relationships between X and G. It will find the absolute value of the correlation matrix between X and G. Then it will sum across all elements of G for each element X and perform a k-means clustering on the absolute correlation sums. All elements of X that correspond to the lowest cluster are noted (not yet filtered unless criteria from the first round is also met). A similar process is again repeated for G by taking the sums across all elements in X for each element in G.

All elements of X that were noted in both the first AND second rounds as belonging in the smallest cluster are filtered out. Additionally, all elements of G that were noted in both the first AND second roungs as belonging to the smallest cluster are filtered out.

**Values**

A list containing the following values:

- *corXY*: A vector containing the absolute correlations between X and Y

- *corGY*: A vector containing the absolute correlations between G and Y

- *xSums*: A vector of the absolute correlation sums across all elements in G for each element in X

- *gSums*: A vector of the absolute correlation sums across all elements in X for each element in G

- *removeX_1*: A vector of the indices corresponding to the elements in X that were noted in the first round as belonging to the smallest cluster

- *removeG_1*: A vector of the indices corresponding to the elements in G that were noted in the first round as belonging to the smallest cluster

- *removeX_2*: A vector of the indices corresponding to the elements in X that were noted in the second round as belonging to the smallest cluster

- *removeG_2*: A vector of the indices corresponding to the elements in G that were noted in the second round as belonging to the smallest cluster

- *removeX_both*: A vector of the indices corresponding to the elements in X that were ultimately filtered due to them intersecting between both the first AND second round

- *removeG_both*: A vector of the indices corresponding to the elements in G that were ultimately filtered due to them intersecting between both the first AND second round

**Examples**

```{r}
#Simulate a system with 200 observations
simSystem <- simPathwaySystem(
  alpha = 0.35
  ,Rsq = 0.35
  ,nsample = 200
  ,rind = 0.35
  ,nx = 15
  ,nxPrime = 100
  ,ns = 50
  ,ntranscriptNoise = 100
  ,ng = 10
  ,ngPrime = 100
  ,nh = 30
  ,ngeneNoise = 100
  ,noiseSD = 0.5
)


#Apply the filters to X and G
filterResults <- SuMOFil(
  x = simSystem$x
  ,g = simSystem$g
  ,y = simSystem$y
  ,numClusters_1 = 3
  ,numClusters_2 = 3
)


#Obtain a finalized filtered X
x_filtered <- simSystem$x[, -filterResults$removeX_both]
#Obtain a finalized filtered X
g_filtered <- simSystem$g[, -filterResults$removeG_both]
```
